---
title: "Final Project"
author: "Aksheytha Chelikavada"
output:
  html_document:
    df_print: paged
  pdf_document: 
    latex_engine: xelatex
  word_document: default
date: "2023-05-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# 1 DATA PREPARATION

## Part 1.1
```{r, message = FALSE}
library(dplyr)
saipe_raw <- read.csv("C:\\Users\\akshe\\Downloads\\SAIPE_04-14-2023.csv")

saipe_mn <- saipe_raw %>% filter(!(Name == "Minnesota")) %>% filter(!(Name == "United States"))%>% select(Year, FIPS = ID, Name, Pop = Poverty.Universe, Poverty = Number.in.Poverty) 
```

Find the largest county, and the nine largest counties by population
```{r}
largest_county_pop <- saipe_mn %>% group_by(FIPS, Name) %>% summarize(Pop = mean(Pop, na.rm = TRUE)) %>%  arrange(desc(Pop)) %>% head(n = 9)
largest_county_pop

FIPSvalue <- saipe_mn %>% group_by(FIPS, Name) %>% summarize(Pop = mean(Pop, na.rm = TRUE)) %>%  arrange(desc(Pop)) %>% head(n = 9) %>% pull(FIPS)

biggest_county <- saipe_mn %>% group_by(FIPS, Name) %>% summarize(Pop = mean(Pop, na.rm = TRUE)) %>%  arrange(desc(Pop)) %>% head(n = 1)
biggest_county
```

Make a time plot showing the number in poverty for each of the nine largest counties
```{r, message = FALSE}
library(dplyr)
library(ggplot2)
library(gtrendsR)
library(tsibble)
library(feasts)

saipe_mn_tsibble <- saipe_mn %>% as_tsibble(index = Year, key = c(FIPS, Name)) %>% filter(FIPS %in% FIPSvalue)

saipe_mn_tsibble  %>% autoplot(Poverty) + facet_wrap(vars(Name))
```


## Part 1.2
```{r, message = FALSE}
library(stringr)
library(lubridate)
library(tidyverse)
library(readr)

cntySnap_raw <- read.csv("C:\\Users\\akshe\\Downloads\\cntysnap.csv",skip = 4, sep ="," )

mnCnty <- cntySnap_raw %>% filter(grepl("MN", Name))

code_mnCnty <- mnCnty %>% mutate(FIPS = paste("27",str_pad(County.FIPS.code, width = 3, pad = "0"), sep = ""))

pivot_code_mnCnty <- code_mnCnty %>% pivot_longer(cols = starts_with("Jul")) %>% mutate(value = as.integer(str_remove(value, ","))) %>% filter(FIPS %in% FIPSvalue) %>% mutate(Year = year(yearmonth(name))) %>% as_tsibble(index = Year, key = c(FIPS, Name))

pivot_code_mnCnty  %>% ggplot2::autoplot((value)) 
```


## Part 1.3
```{r}
raw_irs <- read.csv("C:\\Users\\akshe\\Downloads\\irs.csv", skip = 4)

ts_irs <- raw_irs %>% filter(Name == "Minnesota") %>% mutate(Poor.exemptions = as.integer(str_remove(Poor.exemptions, ","))) %>% as_tsibble(index = Year)

ts_irs %>% autoplot(Poor.exemptions)
```


## Part 1.4
```{r, message = FALSE}
library(lubridate)
library(tidyverse)
library(readr)

join_ts_irs <- raw_irs %>% filter(Name == "Minnesota") %>% dplyr::select(Year, Poor.exemptions) %>% mutate(Poor.exemptions = as.integer(str_remove(Poor.exemptions, ","))) %>% as_tsibble(index = Year)

pivot_code_mnCnty_all <- code_mnCnty %>% pivot_longer(cols = starts_with("Jul")) %>% mutate(value = as.integer(str_remove(value, ","))) %>%  mutate(Year = year(yearmonth(name))) %>% as_tsibble(index = Year, key = c(FIPS, Name))

join_mnCnty_all <- pivot_code_mnCnty_all %>% dplyr::select(FIPS, value, Year)

new_join_mnCnty_all <- join_mnCnty_all %>% mutate(FIPS = as.integer(FIPS))

saipe_mn_join1 <- left_join(saipe_mn, new_join_mnCnty_all, by=c('Year','FIPS'))  

final_join_ts <- left_join(saipe_mn_join1, join_ts_irs, by = 'Year') %>% filter(Year >= 1997) %>% as_tsibble(index = Year, key = c(FIPS, Name.x))

graph_final_ts <- final_join_ts %>% filter(FIPS %in% FIPSvalue)

graph_final_ts %>% autoplot(Pop)
graph_final_ts %>% autoplot(Poverty)
graph_final_ts %>% autoplot(value)
graph_final_ts %>% autoplot(Poor.exemptions)

graph_final_ts %>% as_tibble() %>% ggplot(aes(x = Pop, y = Poverty, color = Name.x)) + geom_point() + facet_wrap(vars(Name.x), scales = "free")
graph_final_ts %>% as_tibble() %>% ggplot(aes(x = value, y = Poverty, color = Name.x)) + geom_point() + facet_wrap(vars(Name.x), scales = "free")
graph_final_ts %>% as_tibble() %>% ggplot(aes(x = Poor.exemptions, y = Poverty, color = Name.x)) + geom_point() + facet_wrap(vars(Name.x), scales = "free")
```


# 2 Linear Models

## Part 2.1
```{r, message = FALSE}
library(forecast)
library(dplyr)
library(lubridate)
library(fpp3)
test_final_ts <-  final_join_ts %>% model(t1 = TSLM(log(Poverty) ~ log(Pop)),
                                           t2 = TSLM(log(Poverty) ~ log(value)),
                                           t3 = TSLM(log(Poverty) ~ log(Poor.exemptions)),
                                           t4 = TSLM(log(Poverty) ~ log(Poor.exemptions)+log(value)),
                                           t5 = TSLM(log(Poverty) ~ log(Poor.exemptions)+log(Pop)),
                                           t6 = TSLM(log(Poverty) ~ log(Pop)+ log(value)),
                                           t7 = TSLM(log(Poverty) ~ log(Pop) + log(value) + log(Poor.exemptions)))


glance(test_final_ts) |> group_by(.model) %>% summarise(CV = sum(CV), AIC = sum(AIC)) %>% arrange(CV, AIC) |>
  dplyr::select(.model,CV, AIC)
```
The model that does the best across all counties is t6 which is TSLM(log(Poverty) ~ log(Pop)+ log(value)). This best model includes poverty, population, and value. 

```{r}
bestModel <- final_join_ts %>% model(TSLM(log(Poverty) ~ log(Pop)+ log(value)))

bestModel %>% filter(FIPS %in% FIPSvalue) %>% augment() %>% autoplot(Poverty) + geom_line(aes(y = .fitted), color = "Black") + facet_wrap(vars(Name.x), scales = "free_y") + theme(legend.position = "none")  
```


## Part 2.2
```{r}
plotRes<- bestModel %>% filter(FIPS %in% FIPSvalue) %>% augment()

autoplot(plotRes, .innov) + facet_wrap(vars(Name.x))

allCountyFIPS <- bestModel %>% pull(FIPS)

bestModel %>% augment() %>% features(.innov, ljung_box) %>% arrange(lb_pvalue)
```
I found one county that was significantly different from white noise. The FIPS code for the county is 27097 and the name is Morrison County.

Because this p-value is so significantly different from white noise, I an going to make a residual plot of it
```{r}
bestModel %>% filter(FIPS == 27097) %>% augment() %>% autoplot(Poverty) + geom_line(aes(y = .fitted), color = "Magenta") + facet_wrap(vars(Name.x), scales = "free") + theme(legend.position = "none") 
```
This model did a pretty good job as we have a few p values below 0.05. Except for one exception, Morrison county got a p value of 0.0009.


# 3 Stochastic Models

## Part 3.1
```{r}
saipe_hen <- saipe_mn %>% filter(!(Year %in% c(1996, 1989, 1993, 1995, 1997))) %>% as_tsibble(index = Year, key = c(FIPS, Name)) %>% filter(FIPS == 27053)

hen_model <-  saipe_hen %>% model(naive = NAIVE(log(Poverty)),
                                           mean = MEAN(log(Poverty)),
                                           ses = ETS(log(Poverty) ~ error("A") + trend("N") + season("N")),
                                           adDamp = ETS(log((Poverty)) ~ error('A') + trend('Ad')),
                                           ad = ETS(log((Poverty)) ~ error('A') + trend('A')),
                                           mul = ETS(log((Poverty)) ~ error('M') + trend('A')),
                                           arima = ARIMA(log(Poverty)))

hen_model  %>% forecast(h = '5 year') %>%  autoplot(saipe_hen) + facet_wrap(.~.model)
```

```{r}
glance(hen_model)
```
The ARIMA model works the best.


## Part 3.2
```{r}
saipe_all <- saipe_mn %>% filter(!(Year %in% c(1996, 1989, 1993, 1995, 1997))) %>% as_tsibble(index = Year, key = c(FIPS, Name)) 

all_model <-  saipe_all %>% model(
                                           ses = ETS(log(Poverty) ~ error("A") + trend("N") + season("N")),
                                           adDamp = ETS(log((Poverty)) ~ error('A') + trend('Ad')),
                                           ad = ETS(log((Poverty)) ~ error('A') + trend('A')),
                                           mul = ETS(log((Poverty)) ~ error('M') + trend('A'))) 


all_model  %>% forecast(h = '5 year') %>%  autoplot(saipe_hen) + facet_wrap(.~.model)

glance(all_model) |> group_by(.model) %>% summarise(AIC = sum(AIC)) %>% arrange(AIC) |>
  dplyr::select(.model, AIC)
```
The ses model did the best compared to the rest of the models. The reason why I chose ses is because it has the lowest AIC score


## Part 3.3
```{r}
arimaFit <- saipe_all %>% model(ARIMA(log(Poverty)))
arimaFit

arimaFits <- saipe_all %>% model(fit100 = ARIMA(log(Poverty) ~ 1 + pdq(1,0,0)),
                                 fit001 = ARIMA(log(Poverty) ~ 1 + pdq(0,0,1)))

glance(arimaFits) |> group_by(.model) %>% summarise(AIC = sum(AIC)) %>% arrange(AIC) |>
  dplyr::select(.model, AIC)
```
(1,0,0) with mean and (0,0,1) with mean are the most common. from the data, (1,0,0) did the best.


## Part 3.4
```{r crossValidate, cache = TRUE}
saipe_all_tr <- saipe_all |>
  stretch_tsibble(.init = 15, .step = 1)

fit_mn <- saipe_all_tr |>
  model(fit100 = ARIMA(log(Poverty) ~ 1 + pdq(1,0,0)),
        ses = ETS(log(Poverty) ~ error("A") + trend("N") + season("N")))


#{r crossValidate, cache = TRUE}
```

```{r}
acc <- fit_mn  %>% forecast(h = 5) %>% fabletools::accuracy(data = saipe_all) 

acc %>% group_by(.model) %>% summarize(sqrt(sum(RMSE*RMSE)))
```
fit100 is the winning model


# 4 Forecasts
```{r}
county_fit <- saipe_all %>% model(fit100 = ARIMA(log(Poverty) ~ 1 + pdq(1,0,0)))

forecast_county_fit <- county_fit %>% forecast(h = '5 year') %>% filter(Year == 2026)
forecast_county_fit

saipe_all_2021 <- saipe_all %>% filter(Year == 2021)

predInterval <- forecast_county_fit$.mean - saipe_all_2021$Poverty

percentInc <- predInterval / saipe_all_2021$Pop
percentInc

highestValues <- tail(sort(percentInc), 5)
index <- which(percentInc %in% highestValues)

fiveCounties <- saipe_all_2021[c(1,4,41,58,63), 'Name']
fiveCounties
```

```{r, message = FALSE}
library(usmap)
library(ggplot2)
forecast_county_fit_usmap <- forecast_county_fit %>% as_tibble()

colnames(forecast_county_fit_usmap)[1] <- "fips"

names(forecast_county_fit_usmap)

plot_usmap(data = forecast_county_fit_usmap, values = ".mean", include = c("MN"), color = "blue") + 
  scale_fill_continuous(low = "white", high = "blue", name = "Poverty Estimates", label = scales::comma) + 
  labs(title = "Minnesota", subtitle = "Poverty Estimates for Minnesota Counties in 2026") +
  theme(legend.position = "right")
```

